Loyalty's Limits: The Impact of Postmodernism and Political Polarization on Military Professionalism  (Prospectus)

### 2020-10-02 code/prospectusv1.Rmd
no message

Figure 1, below, shows a visual diagram of Huntington's logic in a political science framework [@feaver_civil-military_1996]. As Peter Feaver points out in a 1996 article, Huntington did not use such language when developing and explaining his concept. According to Huntington, the ultimate goal, or dependent variable, of civil-military relations in the American context is achieving both military security and maintaining civilian supremacy over the military. Huntington viewed the key independent variable that impacts and enables these goals while also resolving the tension generated by the opposing functional and societal imperatives, as described above, the adoption of a model of "objective civilian control" of civil-military relations, whereby political and military leaders "maximize military professionalism," which includes granting the military a high degree of professional autonomy [@huntington_soldier_1957, 80-85].

@@

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1: Samuel Huntington's Theory\n Of Civilan Control of the Military", subtitle="As espoused in The Soldier and The State, 1957")

@@

Huntington's argument in support of "objective control" of the military stands in contra-distinction, as he explains, to an alternate model of "subjective control" in which civilian leaders limit the power of the military by treating it as any other civilian interest group [@huntington_soldier_1957, 80-83]. In fact, Huntington admits that the US Constitution actually envisioned this "subjective" scheme of civilian control of the military [@huntington_soldier_1957, 163-168]. What the framers did not see, nor could they, argues Huntington, was the rise of the military as a profession in response to ever-increasing and dangerous external threats. It was because of these pernicious threats, argues Huntington, which the US military had witnessed, confronted, and battled during the 19th and 20th centuries during World War One, World War Two, and now the Cold War, that the military profession had risen as a separate and distinct body that could focus on the demands of military service. Such a separation and distinctness, claims Huntington, would need to be embraced by civilians and the militarym going forward into the future. In so doing, claims Huntington, is the only way that both military security and civilian control of the military - two often-competing imperatives - will be safeguarded.

Numerous critiques surrounding the Huntingtonian ideal of _military professionalism_ have been levied over the years, as Huntington did not attempt to precisely define it. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], charging that war by its very nature requires the highest integration - not separation - of military and political activities. In essence, these critiques argue that the strict separation of the responsibilities between the military and civilians leads to poor strategic outcomes [@rapp_civil-military_2015; @brooks_paradoxes_2020].  Moreover, these criticisms can and do point largely to American failures in Iraq and Afghanistan to secure, at least as of yet, a more-lasting peace - failures that are ultimately, according to one scholar and former general, moral in nature, and shared between both civilians and the military [@dubik_just_2016].

As an extension of the critiques of Huntington, there has also been renewed focus and attention among civil-military relations scholars on numerous professional normative standards of the military. One of the most important -  yet difficult - debates is an attempt by scholars to define where appropriate political behavior ends and partisan activity begins [@owens_military_2015; @golby_beyond_2015]. In the Huntingtonian framework, this is essentially the opposite side of the debate over the same coin in attempting to define for civilians what constitutes meddling in military affairs. One of the reasons this debate has become even more difficult, I believe, is because as political polarization increases, the lines between partisanship and politics, as well as politics and everything not politics, have blurred, as so much of the literature on American politics and polarization affirms.

@@

Loyalty features prominently in the international relations and civil-military relations literatures, often in a comparative sense and in the context regime exclusion and a host of related phenomena, including military coups [need cite], civil wars [@mclauchlin_loyalty_2018], and the difficulties that multi-ethnic states face in maintaining a professional military [@peled_question_1998]. This literature deals with a host of important issues that stem from problems associated with ethnic identification and loyalty, either to a regime, cause, or leader. Though the US context is very different from the contexts usually explored in this literature, its insights may help me identify related dynamics and/or mechanisms that perhaps are unfolding in the US: how are changes in partisan or ideological identification (instead of ethnic) as a result of political polarization impacting conceptions of loyalty within the military. Because appropriate to explore partisan or ideological (instead of ethnic) identification stemming from However, this literature often assumes a definition of loyalty and rarely develops how the reader should think of loyalty in the first place. For this fundamental starting point of defining loyalty, the fields of ethics, management, and sociology are helpful.

Loyalty has been conceptualized as both a virtue [@coleman_problems_2009] and as a moral emotion [@connor_military_2019]. Though the differences between the two approaches are important, both assert that loyalty is something felt by an individual and subsequently reflected in action, such as, in the military context, "the desire to protect each other from harm and to defend each other from attack" [@coleman_problems_2009, p. 109]. Oldenquist, a philosopher, describes loyalty as a devotion or attachment that an individual has to someone or something specific such that one refers to the object of loyalty as "mine" and is inclined to feel "pride when it [the object of one's loyalty] prospers, shame when it declines, and anger or indignation when it is harmed" [@oldenquist_loyalties_1982, p. 175, brackets mine].

Loyalty further generates a willingness in the individual to assume "obligations" [@oldenquist_loyalties_1982, p. 175] to the object of one's loyalty, and these obligations can take multiple forms, such as a willingness of the individual to endure hardships and make sacrifices [@fielder_organizational_1992]. Scholars agree that individuals can hold multiple loyalties simultaneously, and recognize that this often can and does cause problems [@coleman_problems_2009]. For organizations, and by extension, professions, the concept of loyalty is also very important. Conner et al. state, for example, that "having loyal members is vital to perpetuating the group" [@connor_military_2019].

We also know that the military services in the United States place a premium on loyalty as a trait and concept. Indeed, of the different military services, the US Army in particular has extensively developed, in a normative sense, a framework entitled "The Army Ethic," in which loyalty constitutes a significant part of the Army profession [@us_department_of_the_army_adp_2019]. This ethic, which will be described in detail in later parts of this dissertation, is formally laid out and described in the Army's doctrinal publication entitled, _Army Leadership and the Profession, ADP 6-22_, and holds that as "stewards of the Army Profession" and as "committed professionals," Army leaders "embrace and uphold the Army Values and standards of the profession" [@us_department_of_the_army_adp_2019]. There are a total of seven "Army Values," which together spell the acronym, l-d-r-s-h-i-p, or "leadership," and the first of these values, loyalty, is followed by six others: duty, respect, selfless service, honor, integrity, and personal courage [@us_department_of_the_army_adp_2019, p. 1-9.].  The US Army states in its doctrinal manual on Army Leadership and the Profession, ADP 6-22, that the Army Values "embody the practical application of the Army Ethic. They encompass the enduring moral principles, beliefs, and laws that guide Army professionals in accomplishing the mission as well as their conduct in all aspects of life" [@us_department_of_the_army_adp_2019, p. 1-12]. The manual defines loyalty as, "loyalty: bear true faith and allegiance to the Constitution of the United States, the Army, your unit and other Soldiers" [@us_department_of_the_army_adp_2019]. The virtue of loyalty is likewise extolled in a host of codes and mantras that have played a central part of the American military since its inception. In the middle of the West Point Cadet Prayer is the eloquently written plea, "make us to choose the harder right instead of the easier wrong" [@wheat_west_1918]. Even Huntington spent a great deal talking about the importance of loyalty as a trait to the very essence of the military profession. He recognized that in order for civilian control to be effective, the military profession had to organize itself as a "hierarchy of obedience," which required "loyalty and obedience" to stand out as "the highest military virtues" [@huntington_soldier_1957, 73].

Loyalty is further important in a military setting in a broader and strategic sense, especially when one considers that the US military, both by statute and by normative standards, serves as an apolitical and non-partisan institutions with multiple, simultaneous loyalties. The Constitution of the United States, to which military officers and soldiers alike ascribe allegiance when they take their respective oaths of commissioning and enlistment, deliberately contrives the military as an institution possessing multiple loyalties. One of the most basic forms of these multiple loyalties is captured in the fact that the Constitution empowers the President as "the Commander in Chief of the Army and the Navy of the United States" [@noauthor_us_1787], but also gives Congress an array of tremendous powers and responsibility pertaining to the Armed Forces @[noauthor_us_1787-1]. Indeed, the tensions that inherently exist in American's constitutional framework are explored in a robust literature that examines the multiple ways in which the Congress, the President, and the military interact to implement, but also to influence, American foreign policy [for instance, see @howell_while_2007].

In the context of the US military as an institution, loyalty has rarely been used as a conceptual construct in the political science literature, likely and thankfully because the US military has not yet considered, or been involved, in a coup. Nonetheless, I contend that there is a rich opportunity to study strategic loyalty as a concept, especially in a contemporary context in the United States, where political polarization has increased extensively.

@@

One area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity of the military, both as an American institution and as a profession. Exploring this relationship will, I believe, help scholars in the field of civil-military relations and international relations more broadly better understand if and how the military, as a unique body charged with the defense of the nation, changes in fundamental ways as a result of a polarized political climate. On one hand, one would expect that it might, in the sense that the American military is, after all, a microcosm of American society. Therefore, one could expect that if society truly is becoming polarized, then at least to some degree, so is the military, or at least those who volunteer for service and at the point in time that they join. On the other hand, one might reasonably expect that the unique purpose of the military, combined with its long-standing professional ethic, might insulate the military institution from some of the potentially negative impacts of polarized politics. These differing expectations need to be more clearly articulated and explored.

I am specifically interested in the question of if and how political polarization impacts the concept of organizational loyalty in the military. This is an interesting research question because, as noted above, one of the most fundamental normative hallmarks of the military's professional nature is that it ought to be apolitical and non-partisan, and one of the features of militaries in general is their tendency to foster a concept of loyalty among those who serve. Yet a puzzle emerges: within an institution that is normatively apolitical and non-partisan, can it remain so in an environment marked by increasingly polarized domestic politics, where partisan loyalties are a salient feature of the political landscape [need cite here for partisan loyalties].

My intended dissertation will help answer several related questions, including: if personal and partisan loyalties are changing as a result of political polarization [need cite], is this also happening within the military institution, and why? Are partisan/ideological loyalties displacing other loyalty relationships within the military? If so, are these changes in organizational loyalty having an impact on the military's professional identity and character? Through what mechanisms do these changes operate? What sorts of changes are they driving, if any, in the military institution? Is this problematic for the military profession, and why?

The answers to these questions, which I intend to include as part of my dissertation, will help inform scholars and policymakers alike on number of important issues. First, it will help illuminate in a broad sense an important aspect of how changes in political polarization impact American institutions, and in particular, the military. Second, work on this topic will attempt to explore how a fundamental concept and value within the military - loyalty - is both impacted as a result of political polarization but also how loyalty changes may be driving changes in the military's professional character, identity, and nature. In doing so, it will help enrich a scholarly understanding of what loyalty is to begin with, and how it features prominently in the context of the American military.

To be clear, scholars have long studied how domestic politics impact the military. And as scholars have pointed out, military and civilian relationships are in tension by their very nature in that the leaders of a democracy allow a most undemocratic institution, the military, to protect the polity [@feaver_civil-military_1996]. And as American history proves, there have been countless instances in which civilian and military leaders disagreed. Many of these rise to the level of common knowledge: Lincoln fired McClellan because he lost confidence in him as a wartime Commander; Truman sacked MacArthur because the latter was insubordinate regarding the future of war in Asia; and Obama and McChrystal clashed about troop levels in Afghanistan. Beyond these episodes in which President and wartime commander locked horns, consider several other instances of "civilians" and "military" clashing. One episode is certainly the way in which the Vietnam war escalated in ways that, at least according to H.R. McMaster, military generals at the time knew were destined to fail [@mcmaster_dereliction_1998]. A more recent example, and one much studied and commented upon by civil-military relations scholars, is the so-called "Revolt of the Generals" that occurred following the early years of the invasion of Iraq, which involved several recently-retired generals and admirals sharply criticizing Secretary of Defense Rumsfeld's leadership dynamics [@cook_revolt_2008].

My personal motivation in choosing the intended topic, however, stems from what I believe are changes in the type of civil-military disagreement that have surfaced in recent years. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, way outside of appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation] More interestingly, these conflicts are occurring not in the context of major war, but the opposite: a period in which the levels of US Troops deployed in active combat zones is less than it has been in recent years. While I believe that such developments are catching the attention of civil-military relations scholars, most of the work done on this topic is, in my opinion, impressionistic [@golby_jim_keep_2020] as opposed to systematic. In other words, civil-military relations scholars have identified that polarized partisan politics serve as a risk for the military, but none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and how.

Thankfully, a growing number of civil-military relations scholars have looked at various aspects of this question - the impact of the political polarization on civil-military relations - in recent years. But the focus of these works has been, I believe, not on the impact of polarization on the military itself but on other important but peripheral components of the military's relationship with American society. One recent dissertation completed by Michael Robinson at Stanford University in 2018, for instance, found that partisan affiliation significantly affects public credibility and trust in the military, perhaps resulting in or as a consequence of media "echo chambers" which may incentivize military leaders to offer opinions partisan opinions without fear of sanction [@robinson_danger_2018]. Robinson's work builds off of previous work by Heidi Urben in which she found that active military officers believe less strongly that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013], a norm that has existed for decades or more. Together, Robinson and Urben join other scholars in providing a wealth of information regarding public opinion and the military, and in particular, how aspects of political polarization shape partisan differences regarding topics such as the use of force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

>>>>

Figure 1, below, shows a visual diagram of Huntington's logic in a political science framework [@feaver_civil-military_1996]. As Peter Feaver points out in a 1996 article, Huntington did not use such language when developing and explaining his concept. According to Huntington, the ultimate goal, or dependent variable, of civil-military relations in the American context is achieving both military security and maintaining civilian supremacy over the military. Huntington viewed the key independent variable that enables the achievement of these goals while also resolving the tension generated by the opposing functional and societal imperatives as the adoption of a type of civil-military relations framework which he calls, "objective civilian control" of the military [@huntington_soldier_1957, 83]. In such a model of "objective civilian control," Huntington envisioned civilians granting the military a high degree of professional autonomy, which he believed would simultaneously be "most conducive to the emergence of professional attitudes and bahvior among members of the officer corps" as well as the best way to minimize the military's political power by rendering the military "politically sterile and neutral" [@huntington_soldier_1957, 83-84], and thus preserving civilian supremacy.

@@

huntington_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MSCS"="Military Security and\n Civilian Supremacy", 
                     "CC"="Patterns of\n Civilian Control\n 'Objective Control'", 
                     "FI"="Functional\n Imperative", 
                     "SI"="Societal\n Imperative",
                     "I"="Liberal Ideology",
                     "S"="Constitutional Structure")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1: Samuel Huntington's Theory Of Civilan Control of the Military", subtitle="As Explained in The Soldier and The State, 1957")

@@

Huntington's argument in support of "objective control" of the military stands in contra-distinction, as he explains, to an alternate model of "subjective control" in which civilian leaders limit the power of the military by treating it as any other civilian interest group [@huntington_soldier_1957, 80-83]. In fact, Huntington admits that the US Constitution actually envisioned and established a form of subjective vice objective scheme of civilian control of the military [@huntington_soldier_1957, 163-168]. What the framers did not see, nor could they, argues Huntington, was the rise of the military as a profession in response to ever-increasing and dangerous external threats. It was because of these pernicious threats, argues Huntington, which the US military had witnessed, confronted, and overcome during the 19th and 20th centuries in World War One, World War Two, and as Huntington writes, the Cold War, as well as the destructive capability of modern weapons and warfare in general that the military profession had emerged [@huntington_soldier_1957]. Huntington claims that unlike during the nation's founding, military officers constitute a profession, and thus, are unique and distinct from other occupations in several important ways: professions practice and develop a special body of knowledge and expertise; professions exercise a  "social responsibility" by providing a vital service to society; and professions have a degree of "corporateness" in that they develop and formalize standards by which members of the profession are certified to practice [@huntington_soldier_1957, 11-18].

Numerous critiques surrounding the Huntingtonian conception of _military professionalism_ have been levied over the years. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres, which Huntington to a certain degree certainly argued for, is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], charging that by its very nature, war requires the highest integration, as opposed to segregation, of political and military activities. Moreover, these critiques often point out, correctly in this author's opinion, that the conduct of war is more complicated than civilians simply telling the military the goals it wishes to achieve, and then allowing the military to determine how best to accomplish these goals. Rather, the development and implementation of strategy, especially during a time of war, requires an "iterative" dialogue between military leaders and civilians to ensure that effective policies are adopted [@rapp_civil-military_2015, 20].

Others argue that Huntington's conception of military professionalism is too narrow to be of practical use in the contemporary period in that such a model actually enables, rather than constrains, inappropriate political activity by military officers [@brooks_paradoxes_2020]. Risa Brooks charges that military officers who strongly identify with an ethic that stresses remaining apolitical at all costs, which she alleges follows from a Huntingtonian conception of military professionalism, may develop "blind spots such that they fail to recognize the political content or impact of their actions" [@brooks_paradoxes_2020, 17]. Whether military leaders recognize it or not, Brooks argues, military leaders comprise a profession that is inherently political, and such a reality ought to be acknowledged and embraced [@brooks_paradoxes_2020].

Regardless of whether and the degree to which one agrees or disagrees with Huntington's concept of military professionalism, a significant number of scholars note that a significant issue involves attempting to define where the line between appropriate political expression and inappropriate partistan activity should be drawn [@owens_military_2015; @golby_beyond_2015; @brooks_paradoxes_2020]. It is a difficult task for many reasons because, as Mackubin Thomas Owens points out, civil-military relations in the United States can be viewed as a "bargain" [@owens_us_2011] between the American people, the military, and the government that has to account for constantly-changing "political, social, technological, or geopolitical changes" [@owens_is_2016]. In other words, the ground is always moving for the actors involved in civil-military relations.

One of these recent changes, I argue, is increasing political polarization that has engulfed not only government elites, but the nation at large. Brooks (2020) points out three challenges posed by the contemporary period to the military's apolitical ethic, all of which are likely related to the growing problem of political polarization: rising acceptance among military officers of behaviors that contradict an apolitical ethos; increasing "pressures and opporunities" for military officers to express partisan sentiment; and increased politicization of the military by government elites [@brooks_paradoxes_2020, 21-23]. Indeed, as political polarization increases, the lines between partisanship and politics, as well as politics and everything not politics, have blurred, as so much of the literature on American politics and polarization affirms, promising that this challenge with respect to the military institution and its members will likely persist for some time.

@@

As a subject or concept, loyalty features prominently in the international relations and civil-military relations literatures, often in a comparative setting and in the context of coups [@marcum_overthrowing_2016; @johnson_squeaky_2018], civil wars [@mclauchlin_loyalty_2018], and the difficulties that multi-ethnic states face in maintaining a professional military [@peled_question_1998]. This literature deals with a host of important issues that stem from problems associated with ethnic identification and loyalty, either to a regime, cause, or leader. Though the US context is very different from the contexts usually explored in this literature, its insights may help me identify related dynamics and/or mechanisms that are unfolding in the US: how are changes in partisan or ideological identification (instead of ethnic) as a result of political polarization impacting conceptions of loyalty within the US military, if at all? To even begin answering this question, one has to first define what loyalty is and how it works. For this fundamental task, the fields of psychology, ethics, management, and sociology are helpful.

Loyalty has been conceptualized as both a virtue [@coleman_problems_2009] and as a moral emotion [@connor_military_2019], the main difference between the two being that the former is a normative aspirational trait, while the latter is a primed behavioral response. For the purposes of my dissertation, these differences are not consequential, because both assert that loyalty is something that is ultimately felt by an individual and subsequently reflected in action, such as, in the military context, "the desire to protect each other from harm and to defend each other from attack" [@coleman_problems_2009, 109]. Oldenquist, a philosopher, further describes loyalty as a devotion or attachment that an individual has to someone or something specific such that one refers to the object of loyalty as "mine" and is inclined to feel "pride when it prospers, shame when it declines, and anger or indignation when it is harmed" [@oldenquist_loyalties_1982, 175]. Loyalty also generates a willingness in the individual to assume "obligations" [@oldenquist_loyalties_1982, p. 175] on behalf of the object of one's loyalty, and these obligations can take multiple forms, such as a willingness of the individual to endure hardships and make sacrifices [@fielder_organizational_1992]. Scholars agree that individuals can hold multiple loyalties simultaneously, and recognize that this often can and does cause problems [@coleman_problems_2009]. For organizations, and by extension, professions, the concept of loyalty is also very important. Conner et al. state, for example, that "having loyal members is vital to perpetuating the group" [@connor_military_2019].

We also know that the military services in the United States place a premium on loyalty as a trait and concept. Indeed, of the different military services, the US Army in particular has extensively developed, in a normative sense, a framework entitled "The Army Ethic," in which loyalty plays a significant role within the Army profession [@us_department_of_the_army_adp_2019]. This ethic, which will be described in detail in later parts of this dissertation, is formally laid out and described in the Army's doctrinal publication entitled, _Army Leadership and the Profession, ADP 6-22_, and holds that as "stewards of the Army profession" and as "committed professionals," Army leaders "embrace and uphold the Army Values and standards of the profession" [@us_department_of_the_army_adp_2019]. There are a total of seven "Army Values," which together spell the acronym, l-d-r-s-h-i-p, which is short for the word "leadership," and the first of these values, loyalty, is followed by six others: duty, respect, selfless service, honor, integrity, and personal courage [@us_department_of_the_army_adp_2019, p. 1-9.].

ADP 6-22 further states that the Army Values "embody the practical application of the Army Ethic. They encompass the enduring moral principles, beliefs, and laws that guide Army professionals in accomplishing the mission as well as their conduct in all aspects of life" [@us_department_of_the_army_adp_2019, p. 1-12]. The manual defines loyalty as, "loyalty: bear true faith and allegiance to the Constitution of the United States, the Army, your unit and other Soldiers" [@us_department_of_the_army_adp_2019]. The virtue of loyalty is likewise extolled in a host of codes and mantras that have played a central part of the American military since its inception. In the middle of the West Point Cadet Prayer is the eloquently written plea, "make us to choose the harder right instead of the easier wrong" [@wheat_west_1918], indicating that to be loyal to what is right is not only preferable but often a difficult endeavor. Even Samuel Huntington recognized that in order for his model of civilian control to be effective, the military profession had to organize itself as a "hierarchy of obedience," which required "loyalty and obedience" to stand out as "the highest military virtues" [@huntington_soldier_1957, 73].

Loyalty is further important in a military setting in a broad and strategic sense, especially when one considers that the US military, both by statute and by normative standards, serves as institution with multiple, simultaneous loyalties. The Constitution of the United States, to which military officers and soldiers alike ascribe allegiance when they take their respective oaths of commissioning and enlistment, contrives the military as an institution possessing multiple loyalties: to the President as "the Commander in Chief of the Army and the Navy of the United States" [@noauthor_us_1787], but also to Congress in that they likewise hold unique powers and responsibilities pertaining to the Armed Forces [@noauthor_us_1787-1]. The tensions that inherently exist in American's constitutional framework with respect to national security are explored in a robust scholarly literature [for instance, see @howell_while_2007; @milner_sailing_2015].

@@

In the context of the US military, to my knowledge, loyalty has rarely been used as a conceptual construct in the political science literature, likely and thankfully because the US military has not yet considered, or been involved, in a coup. Nonetheless, I contend that there is a rich opportunity to study strategic loyalty as a concept, especially in the contemporary context of the United States' domestic political climate, as political polarization has increased extensively.

One potential area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity and character of the military. I am specifically interested in the question of if and how political polarization impacts the concept of loyalty in the military. This is an interesting research question because it brings together at least three relevant strands of literatures which, when combined, generates an interesting set of puzzles and questions. First, one of the most fundamental normative hallmarks of the military's professional nature, at least by Huntingtonian standards, is that it ought to be an apolitical and non-partisan institution. Second, one of the features of the US military is that it prizes and elevates the trait of loyalty among those who serve and, by statute as well as normative tradition, exercises simulataneous dual loyalties to its civilian heads, the President as the Commander in Chief, and Congress. And third, we know from the American politics literature that ideological polarization and partisan loyalty are connected in that those citizens who hold to stronger ideological convictions tend to be the most loyal partisans in terms of voting and other political behaviors [@abramowitz_great_2018, 112-116].

If these things are true, then a question is whether rising political polarization will generate instances in which military officers must choose between which objects of loyalty to ultimately elevate and side with in the event that there is conflict between and among these objects. Moreover, at the level of the military institution, a similar problem may also emerge: will the military institution have to choose between competing objects of loyalty, such as the President as the Commander in Chief, the Congress, and even its own professional ethic? ^ [I believe we have already seen such instances of exactly this situation playing out over the past several years as political polarization has worsened.As examples, consider the case of now retired Army Lieutenant Colonel Alex Vindman, who testified before Congress against the US President, Donald Trump, regarding what be believed to be behavior that may have been illegal by the President regarding aid to Ukraine, or the case of Navy Captain Brett Crozier, who leaked a letter to the press about how his pleas for help in battling the coranavirus aboard the ship he commanded, the USS Theodore Roosevelt. Certainly these instances involved individuals making choices about to whom or to what they would be most loyal to - the President, the Constitution, the ideals and obligations of the military, themselves, and perhaps more.]

Other questions I wish to explore include whether there patterns in the way that loyalty obligations held by the military are in fact challenged by governmental actors during periods of increased political polarization, and if so, what are they, and how is the military reacting to these patterns? Are there impacts on the military's professional identity as a result of political polarization, and if so, from who or what? Are partisan/ideological loyalties displacing other loyalty relationships within the military?  Through what mechanisms do these changes operate? What sorts of changes are they driving, if any, in the military institution? Is this problematic for the military profession, and why? These and other questions will shed light on how rising political polarization is impacting the professional identity of the military.

The gaps in the current literature, which I hope to advance and fill, involve the impact of political polarization directly on the military institution itself, and why the military experiences such impacts. A number of excellent and recent scholarly works have looked at aspects of this question, but from an angle that examines not the military directly, but rather society's perception of the military or various aspects of the military as a result of polarized politics. A number of these works have directly examined the impact of partisan affiliation on the credibility of military actors, resulting in some very interesting findings. Michael Robinson, for instance, found that members of the public did not sanction military elites for making public messages that accorded with a member's partisan views, and that the public only sanctioned those elites whose messages were in line with the opposite political party [@robinson_danger_2018]. Previously, Heidi Urben found that among active military officers, there was weakening support for the norm that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013]. These scholars' findings suggest, as others have pointed out [@brooks_paradoxes_2020], that political polarization at least coincides with, if not causes, a weakening in support among both the public and among service members for the military's normative apolitical ethos. Other scholars have also unpacked a trove of information regarding public opinion and the military, including how partisan affiliation shapes differences in support for the use of military force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

Focusing attention on the military institution will help scholars in the field of civil-military relations and international relations more broadly better understand if and how the military, as a unique body charged with the defense of the nation, changes in fundamental ways as a result of a polarized political climate. On one hand, one would expect that it might, in the sense that the American military is, after all, a microcosm of American society. Therefore, one could expect that if society truly is becoming polarized, then at least to some degree, so is the military, or at least those who volunteer for service and at the point in time that they join. On the other hand, one might reasonably expect that given the unique purpose of the military, combined with its long-standing professional ethic, might insulate the military institution from some of the potentially negative impacts of polarized politics. These differing expectations need to be more clearly articulated and explored.

To be clear, scholars have long studied how domestic politics impact the military. And as scholars have pointed out, military and civilian relationships are in tension by their very nature in that the leaders of a democracy allow a most undemocratic institution, the military, to protect the polity [@feaver_civil-military_1996]. And as American history proves, there have been countless instances in which civilian and military leaders disagreed. Many of these rise to the level of common knowledge: Lincoln fired McClellan because he lost confidence in him as a wartime Commander; Truman sacked MacArthur because the latter was insubordinate regarding the future of war in Asia; and Obama and McChrystal clashed about troop levels in Afghanistan, ultimately resulting in the former firing the latter. Beyond these episodes in which President and wartime commander locked horns, consider several other instances of "civilians" and "military" clashing. One episode is certainly the way in which the Vietnam war escalated in ways that, at least according to H.R. McMaster, military generals at the time knew were destined to fail [@mcmaster_dereliction_1998]. A more recent example, and one much studied and commented upon by civil-military relations scholars, is the so-called "Revolt of the Generals" that occurred following the early years of the invasion of Iraq, which involved several recently-retired generals and admirals sharply criticizing Secretary of Defense Rumsfeld's leadership dynamics [@cook_revolt_2008].

My intended topic is different than these other, still very important topics in which disagreements between the executive and the military occur during wartime. It is my assessment that polarization is causing conflict, or certainly has the potential to do so, between the military and civilian leaders, but in most instances, the subject is about topics other than the use of military force, which is somewhat puzzling, given the military's reason d'etre. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, outside the bounds of what constitutes appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation] While I believe that such developments are catching the attention of civil-military relations scholars, none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and how.
